---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("sf")
install.packages("giscoR")
install.packages("tinytex")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
library(readxl)
library(sf)
library(giscoR)
library(dplyr)
library(ggplot2)
library(tinytex)
tinytex::install_tinytex(force = TRUE)
```

# Title Page

Include your names
Noel Kona
Giorgia Tanzi 
Iann Soffried 
Julian Trinkl 
Morris Nouwens 
Dania Al Khulaifi 

Include the tutorial group number
Tutorial 5 Group 5 

Include your tutorial lecturer's name
Simon Loop

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}

rent<-read_xlsx("prc_hicp_aind__custom_12680989_spreadsheet (1).xlsx")

europe <- gisco_get_countries(
  resolution = "20"
)

wage<-read_xlsx("tps00155_page_spreadsheet.xlsx")

```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(wage)

head(rent)

head(europe)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning, show=F}




# =============================================================================
#rent cleaning, deleted NA columns and only selected values after 2014 and then calculated var between 2015 and 2025
# =============================================================================
colnames(rent)<-rent[8,]

rent<-rent[-1:-10,]

rent<-rent[-36:-42,]

rent<-rent[,-seq(from=3, to=ncol(rent), by=2)]

rent<-rent[-33,]

rent<-rent[,-2:-6]

rent<-rent[-34,]

rent <- rent %>%
  mutate(
    `2015` = as.numeric(gsub("[^0-9.]", "", `2015`)),
    `2025` = as.numeric(gsub("[^0-9.]", "", `2025`)),
    Var = (`2025` - `2015`) / `2015`
  )

rent <- rent %>%
  mutate(across(2:ncol(rent), as.numeric))


# =============================================================================
# europe cleaning
#==============================================================================

europe <- europe[europe$CNTR_ID %in% c(
  "NL","DE","FR","BE","IT","ES","PT",
  "AT","PL","CZ","SK","HU","RO",
  "BG","HR","SI","DK","SE","FI",
  "EE","LV","LT","IE","LU","GR",
  "CY","MT","NO","CH","ME","MK","TR","IS","RS"
), ]


# =============================================================================
# wage cleaning
#==============================================================================
colnames(wage)<-wage[7,]

wage<-wage[c(-1:-12,-16,-24:-25,-32,-38:-44,-53:-57),]

wage<-wage[,-seq(from=3, to=ncol(wage), by=2)]

wage<-wage[,-seq(from=3, to=ncol(wage), by=2)]


```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1, show=F}


data<-left_join(europe, rent, join_by("NAME_ENGL"=="TIME"))

```

Variable 2

```{r gen_var2, show=F}



hicp_long <- rent %>%
  pivot_longer(
    cols = matches("^20"),
    names_to = "year",
    values_to = "hicp"
  ) %>%
  mutate(year = as.numeric(year))

wage_long <- wage %>%
  pivot_longer(
    cols = matches("^20"),
    names_to = "year",
    values_to = "wage_growth"
  ) %>%
 mutate(
  year = as.numeric(str_extract(year, "\\d{4}")),
  wage_growth = as.numeric(wage_growth)
)
analysis <- left_join(
  hicp_long,
  wage_long,
  by = c("TIME", "year")
)

```

## 3.3 Visualize temporal variation

```{r, show=F}

# =============================================================================
# Calculate EU average each year
# =============================================================================
temporal <- analysis %>%
  group_by(year) %>%
  summarise(
    avg_hicp = mean(hicp, na.rm = TRUE),
    avg_wage = mean(wage_growth, na.rm = TRUE)
  ) %>%
  arrange(year) %>%
  mutate(
    avg_hicp        = avg_hicp / avg_hicp[year == 2020] * 100,
    avg_wage_growth = avg_wage / avg_wage[year == 2020] * 100
  ) %>%
  select(year, avg_hicp, avg_wage_growth)




# =============================================================================
# Convert to long format for plotting
# =============================================================================
temporal_long <- temporal %>%
  filter(year >= 2020) %>%
  pivot_longer(
    cols = c(avg_hicp, avg_wage_growth),
    names_to = "variable",
    values_to = "value"
  )




# =============================================================================
# Plot
# =============================================================================
ggplot(temporal_long,
       aes(x = year,
           y = value,
           colour = variable)) +
 geom_line(linewidth = 1.2, na.rm = TRUE) +
geom_point(size = 2, na.rm = TRUE) +
  labs(
    title = "Temporal Variation of HICP and Minimum Wage Growth",
    x = "Year",
   y = "Index (2020 = 100)",
    colour = "Indicator"
  ) +
  scale_colour_manual(
    values = c(
      "avg_hicp" = "red",
      "avg_wage_growth" = "blue"
    ),
    labels = c(
      "HICP Inflation",
      "Minimum Wage Growth"
    )
  ) +
  theme_minimal()

```

## 3.4 Visualize spatial variation

```{r visualise_map, show=F}

map_data <- left_join(europe, rent, join_by("NAME_ENGL" == "TIME")) %>% 
  st_as_sf()

ggplot(map_data) +
  geom_sf(aes(fill = Var)) +
  coord_sf(
    xlim = c(-25, 45),
    ylim = c(34, 72),
    expand = FALSE
  ) +
  labs(title = "Map rent", fill = "Variation") +
  scale_fill_viridis_c(labels = scales::label_percent(accuracy = 1)) +
  theme_void()


```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations, show=F}
# Step 1: which countries have a minimum wage? (the ones that show up in the wage data)
wage_countries <- unique(wage_long$TIME)

# Step 2: label each country as being in one group or the other
subpop <- rent %>%
  mutate(min_wage_status = ifelse(TIME %in% wage_countries,
                                  "Has minimum wage", "No minimum wage"))

# Step 3: boxplot comparing rent inflation between the two groups
ggplot(subpop, aes(x = min_wage_status, y = Var, fill = min_wage_status)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(title = "Rent inflation 2015–2025, by minimum-wage status",
       x = NULL, y = "Change since 2015") +
  theme_minimal() +
  theme(legend.position = "none")
```


Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis, show=f}

# Summarise average wage growth per country over the full period
wage_avg <- wage_long %>%
  group_by(TIME) %>%
  summarise(avg_wage_growth = mean(wage_growth, na.rm = TRUE))

# Join with rent variation
scatter_data <- left_join(
  rent %>% select(TIME, Var),
  wage_avg,
  by = "TIME"
) %>%
  filter(!is.na(Var), !is.na(avg_wage_growth))

# Scatter plot
ggplot(scatter_data, aes(x = avg_wage_growth, y = Var)) +
  geom_point(colour = "steelblue", size = 3) +
  geom_smooth(method = "lm", se = TRUE, colour = "red", fill = "pink") +
  geom_text(aes(label=TIME),vjust = -0.7, size = 2.8) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  scale_x_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(
    title = "Rent Inflation vs. Minimum Wage Growth (2015–2025)",
    x = "Average Minimum Wage Growth (2015–2025)",
    y = "Rent Inflation (2015–2025)",
    caption = "Each point represents a country"
  ) +
  theme_minimal()

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
