---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: 
 pdf_document:
   latex_engine: xelatex
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("sf")
install.packages("giscoR")
install.packages("tinytex")
```

```{r packages, include=FALSE}
require(tidyverse)
require(rmarkdown)
require(yaml)
library(readxl)
library(sf)
library(giscoR)
library(dplyr)
library(ggplot2)
library(tinytex)
```

# Title Page

Noel Kona,
Giorgia Tanzi, 
Iann Soffried, 
Julian Trinkl, 
Morris Nouwens and 
Dania Al Khulaifi 


# Part 1 - Identify a Social Problem

The social problem being discussed is whether the growth of wages keeps up with the rising costs of living. The central question of this research is: does the growth of wages of the lowest-paid workers rise enough for them to be able to afford their current lifestyle? And does that apply across Europe?

## 1.1 Describe the Social Problem

-   The social problem:
The social problem is about the prices of living rising faster compared to wages, resulting in a worker's real purchasing power falling even if their wages rise on paper. This leads to workers being able to afford less each year.

-   The relevance:
The problem of the growing wages and the rising cost of living is relevant, because in recent years many countries in Europe have faced a sharp rise in prices, while the growth of wages falls behind, leaving many households with low-income struggling to financially keep up with the new living costs (ILO, 2022). The EU also implemented Directive (EU) 2022/2041 on adequate minimum wages in 2022, this aims to keep wage levels keep up with the growing inflation and living costs (European Parliament and Council of European Union, 2022).

-   What the report will answer:
The report will fill in the gap by comparing consumer-price inflation and minimum-wage levels across European countries throughout recent years. Existing reports look globally/nationally, but cross-country researches are under-examined, which is why this report will use cross-country methods to deepen such

# Part 2 - Data Sourcing
2.1: Data sourcing 
Eurostat was selected as the data source for both European Union's Harmonized Index of Consumer Prices (HICP) and the European Union's minimum wage. Eurostat uses a harmonised and standardised way of collecting statistics which allow for reliable cross-country comparisons. It collects and compiles data from national statistical offices using common methods to improve reliability and consistency. This makes Eurostat very suitable for determining whether minimum wages relative to the cost of living have been increasing at an adequate rate throughout Europe.
There are some limitations to our data that deserve mention. The HICP data only takes into account average price movements of all consumer goods in a country and does not provide data for regional price variations or for the individual consumer's experience. Furthermore, the minimum wage database only contains data from countries that have followed a statutory National Minimum Wage system. This limits the comparisons between countries that have relied primarily on collective bargaining systems. Lastly, there were missing values for some observations, which resulted in an inability to include several countries in the comparison.

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data, include=FALSE}

rent<-read_xlsx("prc_hicp_aind__custom_12680989_spreadsheet (1).xlsx")

europe <- gisco_get_countries(
  resolution = "20"
)

wage<-read_xlsx("tps00155_page_spreadsheet.xlsx")

```


## 2.2 Provide a short summary of the dataset(s)

```{r, echo=FALSE}
head(wage)

head(rent)

head(europe)
```



This study's aim is to assess wages' rise in line with the country's rising living costs across Europe through the use of three datasets.

The Harmonised Index of Consumer Prices (HICP) was used for tracking rent/housing rates in Europe. The data were sourced from Eurostat and cover the annual rate of inflation for Europe up to the present (2022). The dataset also enables us to analyze the way that housing costs change over time (i.e., rent inflation rates from 2015 – 2025).

The second dataset used is based on the growth of the minimum wage rate in European countries as supplied by Eurostat. The statistical measures capture the annual increase in the statutory minimum wage for the countries in Europe that have a National Minimum Wage. This data allows for comparison of wage increases with others and helps determine if wage growth has been enough to keep up with price increases.

The third dataset used contains the geographical boundaries for European Countries that were supplied through the spatial package 'giscoR'. These geographical boundaries are used to illustrate rent inflation differences between the countries of Europe with maps.

When combined, these three individual datasets allow for comparison of how, both over time the cost to rent housing and minimum wage are related.

## 2.3 Describe the type of variables included

Think of things like:

*-   Do the variables contain health information or SES information?*

*-   Have they been measured by interviewing individuals or is the data coming from administrative sources?*

Socioeconomic and economic indices, measured at the nation level, are the primary components of the datasets.

The rent data set comprises quantitative variables representing HICP (harmonised indices of consumer prices) values per year, which measure the change in consumer prices for housing-related items. This information allows analysts to assess how living costs have changed and how inflation is affecting households.

Wage datasets consist of quantitative variables that indicate how statutory minimum wage rates have changed from one year to another across the various countries. Labour force conditions and income levels among low-wage employees can therefore be analysed based on the available wage indices.

Country boundaries or country identifiers constitute the variables in the geographic dataset. While these geographic variables do not directly measure social or economic conditions, they allow for the geographical visualisation of socioeconomic indices.

The datasets originate from governmental administrations and official statistics rather than personal surveys or interviews. Economic data were gathered and developed by national statistical agencies and published by Eurostat. Thus, the datasets include standardised and equivalent figures across all European countries.
we removed bosnia, turkey, and albania because some data was missing from them.

```{r, echo=FALSE}
variables <- data.frame(
  Variable = c(
    "TIME",
    "year",
    "hicp",
    "geometry"
  ),
  Type = c(
    "character",
    "integer",
    "numeric",
    "spatial"
  ),
  Meaning = c(
    "European country",
    "Year of observation",
    "Harmonised Index of Consumer Prices",
    "Country geographic boundaries"
  )
)

knitr::kable(variables)
```

# Part 3 - Quantifying

## 3.1 Data cleaning

How did we clean the data? 

We processed the HICP, minimum wage, and geographic dataset. This involved a three-step procedure:
1. Cleaning of the data by removing metadata, unnecessary data rows, and unnecessary data columns: The original Eurostat datasets included descriptive metadata, empty or missing values, and variable definitions that were not needed for our research question; we did not require these records, so they were eliminated. Therefore, we retained only those country-level observations and years that were relevant to our analysis.
2. Standardizing and transforming the data into a more consistent format: We changed the format of both HICP and minimum wage values to a numeric format, and restructured each dataset in a consistent country-year format; Additionally, we also created a new variable that reflects the percent change in HICP between 2015 and 2025. This will help in comparing inflation trends between countries.
3. Merging datasets and removing incomplete observations: The cleaned HICP and minimum wage datasets were merged based on country and year matching records; In addition, the geographic dataset was merged based on matching records using country names. Any countries that did not have sufficient records to perform consistent and comparable analyses or visualizations were excluded.

```{r data_cleaning, include=FALSE}



#rent cleaning, deleted NA columns and only selected values after 2014 and then calculated var between 2015 and 2025

colnames(rent)<-rent[8,]

rent<-rent[-1:-10,]

rent<-rent[-36:-42,]

rent<-rent[,-seq(from=3, to=ncol(rent), by=2)]

rent<-rent[-33,]

rent<-rent[,-2:-6]

rent<-rent[-34,]

rent <- rent %>%
  mutate(
    `2015` = as.numeric(gsub("[^0-9.]", "", `2015`)),
    `2025` = as.numeric(gsub("[^0-9.]", "", `2025`)),
    Var = ((`2025` - `2015`) / `2015`)*100
  )

rent <- rent %>%
  mutate(across(2:ncol(rent), as.numeric))



# europe cleaning


europe <- europe[europe$CNTR_ID %in% c(
  "NL","DE","FR","BE","IT","ES","PT",
  "AT","PL","CZ","SK","HU","RO",
  "BG","HR","SI","DK","SE","FI",
  "EE","LV","LT","IE","LU","GR",
  "CY","MT","NO","CH","ME","MK","TR","IS","RS"
), ]

data<-left_join(europe, rent, join_by("NAME_ENGL"=="TIME"))

# wage cleaning

colnames(wage)<-wage[7,]

wage<-wage[c(-1:-12,-16,-24:-25,-32,-38:-44,-53:-57),]

wage<-wage[,-seq(from=3, to=ncol(wage), by=2)]

wage<-wage[,-seq(from=3, to=ncol(wage), by=2)]

# analysis

hicp_long <- rent %>%
  pivot_longer(
    cols = matches("^20"),
    names_to = "year",
    values_to = "hicp"
  ) %>%
  mutate(year = as.numeric(year))

wage_long <- wage %>%
  pivot_longer(
    cols = matches("^20"),
    names_to = "year",
    values_to = "wage_growth"
  ) %>%
 mutate(
  year = as.numeric(str_extract(year, "\\d{4}")),
  wage_growth = as.numeric(wage_growth)
)

analysis <- left_join(
  hicp_long,
  wage_long,
  by = c("TIME", "year")
)

```


## 3.2 Generate necessary variables
## Creating new variables

```{r, echo=FALSE}

# Create percentage change in HICP between 2015 and 2025
rent <- rent %>%
  mutate(
    Var = ((`2025` - `2015`) / `2015`) * 100 )


country_variation <- analysis %>%
  filter(TIME %in% c("Netherlands", "Germany", "France")) %>% 
  group_by(TIME) %>%
  summarise(
    avg_wage_growth = mean(wage_growth, na.rm = TRUE)
  ) %>%
  left_join(
    rent %>%
      select(TIME, Var),
    by = "TIME"
  )

knitr::kable(country_variation)


```

## 3.3 Visualize temporal variation

```{r, echo=FALSE}


# Calculate EU average each year

temporal <- analysis %>%
  group_by(year) %>%
  summarise(
    avg_hicp = mean(hicp, na.rm = TRUE),
    avg_wage = mean(wage_growth, na.rm = TRUE)
  ) %>%
  arrange(year) %>%
  mutate(
    avg_hicp        = avg_hicp / avg_hicp[year == 2020] * 100,
    avg_wage_growth = avg_wage / avg_wage[year == 2020] * 100
  ) %>%
  select(year, avg_hicp, avg_wage_growth)





# Convert to long format for plotting

temporal_long <- temporal %>%
  filter(year >= 2020) %>%
  pivot_longer(
    cols = c(avg_hicp, avg_wage_growth),
    names_to = "variable",
    values_to = "value"
  )





# Plot

ggplot(temporal_long,
       aes(x = year,
           y = value,
           colour = variable)) +
 geom_line(linewidth = 1.2, na.rm = TRUE) +
geom_point(size = 2, na.rm = TRUE) +
  labs(
    title = "Temporal Variation of HICP and Minimum Wage Growth",
    x = "Year",
   y = "Index (2020 = 100)",
    colour = "Indicator"
  ) +
  scale_colour_manual(
    values = c(
      "avg_hicp" = "red",
      "avg_wage_growth" = "blue"
    ),
    labels = c(
      "HICP Inflation",
      "Minimum Wage Growth"
    )
  ) +
  scale_y_continuous(
  limits = c(95, 130)
) +
  theme_minimal()

```
The graph above directly tests our social question. It shows that in 2020 to 2025 the minimum wage growth is growing faster (blue) than the HICP inflation (red). This graph suggests that at the average-level the minimum-wage policy broadly kept up with the cost of living over this time period. Furthermore, it shows that there was most likely a deliberate choice to increase wages after the inflation spikes in 2021-2022. However, keep in mind that this graph shows an average which hides large differences between countries (see sections 3.4 and 3.5).

## 3.4 Visualize spatial variation

```{r visualise_map, echo=FALSE}

map_data <- left_join(europe, rent, join_by("NAME_ENGL" == "TIME")) %>% 
  st_as_sf()

ggplot(map_data) +
  geom_sf(aes(fill = Var)) +
  coord_sf(
    xlim = c(-25, 45),
    ylim = c(34, 72),
    expand = FALSE
  ) +
  
  scale_fill_gradient(
  low = "yellow",
  high = "red",
  labels = scales::label_number(suffix = "%")
) +
labs(
  title = "Map increase in prices between 2015-2025",
  fill = "Increase in rent(%)"
) +
  
  
  
   #scale_color_gradient(low="lightblue", high= "darkblue") +
  #labs(title = "Map increase in prices between 2015-2025", fill = "Variation of rent prices") +
 
  #scale_fill_viridis_c(labels = scales::label_percent(accuracy = 1)) + 
 
  theme_void()


```

The map shows how much consumer-prices rose in each country between 2015-2025 (not each countries data starts in 2015), and makes clear that not each country has the same growth in prices. With some countries having far steeper growths than others. The largest increases are seen in Eastern-Europe, with a clear example being Hungary that had a growth of 100%. Meanwhile, the countries in Western-Europe, such as France, Germany and Italy saw smaller rises (around 25%). This matters for the social problem, because the countries who face far higher growth in costs of living, will require bigger rises in wage otherwise it will not be sufficient and would leave them off worse. This graph also explains that section 3.3 hides this data, this proves that a single trend line cannot capture each country's situation. (Türkiye is shown in gray since there was no data)

## 3.5 Visualize sub-population variation


```{r visualise_subpopulations, echo=FALSE}
# Step 1: which countries have a minimum wage? (the ones that show up in the wage data)
wage_countries <- unique(wage_long$TIME)

# Step 2: label each country as being in one group or the other
subpop <- rent %>%
  mutate(min_wage_status = ifelse(TIME %in% wage_countries,
                                  "Has minimum wage", "No minimum wage"))

# Step 3: boxplot comparing rent inflation between the two groups
ggplot(subpop, aes(x = min_wage_status, y = Var, fill = min_wage_status)) +
  geom_boxplot() +
  scale_y_continuous() +
  labs(title = "Rent inflation 2015–2025, by minimum-wage status",
       x = NULL, y = "Change since 2015") +
  theme_minimal() +
  theme(legend.position = "none")
```


This plot above shows price inflation between countries who have a statutory minimum wage (salmon) and the ones that do not (cyan). Countries with a minimum wage saw both higher and far more variable inflation, a median around 48% and a spread of a 100%, while countries without a minimum wage were lower and tightly clustered together, with a median of 26%. First impression, this looks like having a minimum wage comes with higher inflation, but that is not entirely true. The countries with "no minimum wage" is made up of countries that are wealthy and economically stable (such as Denmark, Sweden, Finland, Austria and Switzerland). Meanwhile the group with minimum wage includes countries from Central Europe and Eastern Europe that have experienced sharp rises in prices. The two group have many differences, not only in minimum wage policies, so this chart shows association, not a cause. For the social problem at hand, it highlights that low-income workers in minimum wage countries often faced far steeper changes in cost-of-living, making it far more important whether their wages kept up.

Two countries in the "no minimum wage" group are outliers, which are Iceland and Austria, showing higher inflation regardless. This shows that the grouping is not always an indicator of whether or not you have high inflation.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis, echo=FALSE}



temporal_long2 <- temporal_long %>%
  # 1. Group by the indicator so calculations don't cross-contaminate
  group_by(variable) %>%
  
  # 2. Ensure the data is in chronological order before calculating
  arrange(year, .by_group = TRUE) %>%
  
  # 3. Calculate the year-over-year percentage change
  mutate(
    pct_growth_yoy = ((value - lag(value)) / lag(value)) * 100
  ) %>%
  
  # 4. Always ungroup when you are done to prevent issues later
  ungroup()


temporal_long2<-temporal_long2[-1,]

temporal_long2<-temporal_long2[-6,]

ggplot(temporal_long2,
       aes(x = year,
           y = pct_growth_yoy,
           colour = variable)) +
 geom_line(linewidth = 1.2, na.rm = TRUE) +
geom_point(size = 2, na.rm = TRUE) +
  geom_vline(xintercept = 2022, 
             linetype = "dotted", 
             color = "darkgray", 
             linewidth = 0.8) +
  scale_y_continuous(limits = c(0,12)) +
  
  # 2. Add the text label just above or near the top of the line
  annotate("text", 
           x = 2022, 
           y = 12,
           label = "Start of Russia War") +
  labs(
    title = "Annual percentual change of HICP and Minimum Wage Growth",
    x = "Year",
   y = "Percentage variation (yoy)",
    colour = "Indicator"
  ) +
  scale_colour_manual(
    values = c(
      "avg_hicp" = "orange",
      "avg_wage_growth" = "purple"
    ),
    labels = c(
      "HICP Inflation",
      "Minimum Wage Growth"
    )
  ) +
  theme_minimal()

```
This graph depicts the yearly difference between HICP inflation and Minimum Wage growth over time. The comparison of the two lines determines whether wages were keeping up with inflation based on changes to the prices of goods and services bought by consumers. For example, prior to 2022, both variables were relatively stable over time and only changed from year-to-year by small amounts. However, due to the start of the conflict between Russia and Ukraine in 2022, inflation rates increased significantly due to increases to energy prices (e.g., electricity, heating) and food prices across Europe. As a result of these inflationary pressures, minimum wage rates also increased as many governments raised their statutory minimum wages in an effort to help their citizens cope with a higher cost of living. 
However, if the inflation line rises above the minimum wage line, this indicates that inflation is increasing faster than the rate of minimum wage increases, thereby reducing the purchasing power of minimum wage workers. Thus, if minimum wage growth is less than or equal to the rate of inflation, then a minimum wage worker’s ability to maintain his/her living standards will be greater than had minimum wage increases surpassed inflation. The significance of this graph to the investigation into minimum wages and inflation is that it shows the clear relationship between cost of living increases and wage increases; specifically, it indicates that although minimum wages may increase in response to and after significant inflationary shocks, the amount of time and/or the extent of minimum wage increases in reaction to inflation may not be enough to offset the increasing cost of living faced by minimum wage earners.
Here you provide a description of why the plot above is relevant to your specific social problem.
#===============================================================================================================================================================================================
# Part 4 - Discussion

## 4.1 Discuss your findings 
The objective of this research was to determine if there was a disparity between the growth of minimum wage and the increases in the cost of living, measured in terms of the European average for all Member States. The analysis found that European averages for minimum wage growth were either at or above European average consumer price growth rates from 2020 through 2025. This is consistent with more recent European policy goals to provide minimum wages that allow for adequate living standards, while also supporting vulnerable or low-income workers to afford necessities in an environment of rising living costs. The Jacques Delors Centre indicates that providing fair and adequate minimum wages is key to long-term provision of living standards and decreasing in-work poverty, especially during times of rapid inflation (Andor, 2022). Likewise, Eurofound reports that increases in the statutory minimum wage provide increased earnings for low-income workers and help offset some of the effects of rising prices.
In addition, although findings from the study indicated countries with higher than average minimum wage growth and countries with higher than average consumer price growth, the study also highlighted the significant geographic variability between countries within the region. The spatial analysis of the countries suggested that inflation occurred at significantly elevated levels in many of the Eastern European countries when compared to many of the Western European countries. The results from this analysis provide evidence to support studies that have reported inflation in Central and Eastern Europe remains significantly higher than the European average due to geographic differences in energy use, economic makeup, and market conditions. Consequently, workers in different countries faced very different cost-of-living pressures even when average European trends appeared favourable.

The analysis of the data indicates that there were higher minimum wage increases due to the inflation shock that arose following the start of the Russia-Ukraine war in February 2022. Past studies show that the war led to a drastic increase in food and energy prices across Europe, causing inflationary pressures on households. Many governments raised their minimum wages as an adjustment to offset these higher prices. However, the results of this study do not show that the war actually caused the increase in wages or that the increase in minimum wage compensated for inflation completely. Instead, what this data shows is that the two events occurred during the same time frame, and they seem to be associated with one another.

Additionally, these findings also indicate that there needs to be a focus on total purchasing power when looking at minimum wage growth. Many countries raised their minimum wages; however, the effects of inflation will impact households differently based on their spending habits. According to research done by De Nederlandsche Bank, low-income households tend to use a greater percentage of their income on necessary items, like food and energy. Therefore, they bear a larger burden of inflation than higher-income families. ABN AMRO also notes that even though average household financial situations have improved, many families still deal with tremendous cost-of-living increases. Therefore, just because a country raises its minimum wage does not mean that all workers experience an improvement in their real standard of living.

It is important to be aware of a few limitations. To begin with, the analysis only looks at the national average for countries, so it will not account for potential differences between areas (i.e., urban vs rural). Housing prices and consumer price indices can differ greatly from one area to another, so this means that the experience of a minimum wage earner will be different than the average experience in the country. Secondly, the minimum wage data collected was only for countries that have minimum wage laws; therefore, countries such as Denmark, Sweden, and Finland cannot be compared as easily due to their having a system based on collective bargaining. As thirdly stated, HICP is a composite index of consumer prices and may not adequately reflect the actual consumption pattern of the low-income population, which is the group most impacted by the minimum wage law. Therefore, this study found associations between increases in inflation and increases in the minimum wage, but it does not provide the ability to measure a cause-and-effect relationship. In addition, factors such as improved productivity, labour market conditions, government intervention, and external economic issues often influence wage growth and inflation.

The data indicate that minimum wages in Europe have increased as living costs rise, helping to safeguard low-income employees from the impact of inflation. However, there is a regional disparity in this protection, and wage increases often do not keep pace with price increases. Future research may focus on specific countries or utilise individual-level household data sets to analyse how wage changes and inflation affect the living standards of different demographic groups.

# Part 5 - Reproducibility

## 5.1 Github repository link


Provide the link to your PUBLIC repository here: https://github.com/iannzito/ProgrammingForEcon
#=============================================================================================================================================================
## 5.2 Reference list

International Labour Organization. (2022). Global Wage Report 2022–23: The impact of inflation and COVID-19 on wages and purchasing power. https://www.ilo.org/publications/flagship-reports/global-wage-report-2022-23-impact-inflation-and-covid-19-wages-and

European Parliament and Council of the European Union. (2022). Directive (EU) 2022/2041 of 19 October 2022 on adequate minimum wages in the European Union. Official Journal of the European Union. https://eur-lex.europa.eu/eli/dir/2022/2041/oj/eng

Cobain, D. (2026, June 5). Minimum Wage in Europe 2026: 27 EU Country Guide. Employsome. https://employsome.com/blog/minimum-wage-europe/

Cova, J. (n.d.). The EU directive on fair and adequate minimum wages: Low-pay work, collective bargaining and the prospects for income growth. Jacques Delors Centre. https://www.delorscentre.eu/en/publications/fair-and-adequate-minimum-wages

De Nederlandsche Bank. (2025, August 25). How inflation affects households differently. www.dnb.nl. https://www.dnb.nl/en/general-news/background-2025/how-inflation-affects-households-differently/

Delbeke, J., Cornillie, J., & Vis, P. (2022, March 24). The impact of the war in Ukraine on Europe’s climate and energy policy. European University Institute. https://www.eui.eu/news-hub?id=the-impact-of-the-war-in-ukraine-on-europes-climate-and-energy-policy

Half of EU countries exclude workers from minimum wage. (2021, June 9). https://www.etuc.org/en/pressrelease/half-eu-countries-exclude-workers-minimum-wage

López-Bazo, E., Ramos, R., Royuela, V., Molina, O., Vacas-Soriano, C., & de Miguel, P. S. (2025). Impact of national minimum wages on collective bargaining and wages for low-paid workers. Europa.eu, 110. Eurofound. https://doi.org/10.2806/2659787

Times, C. E. (2026, February 10). Regional inflation: still standing out above the European line. Central European Times. https://centraleuropeantimes.com/regional-inflation-still-standing-out-above-the-european-line/

van Reeken-van Wee, J., & van de Kerke, J.-P. (2025, October 22). Cost of living improved despite more expensive groceries. ABN AMRO Bank. https://www.abnamro.com/research/en/our-research/transaction-trends-cost-of-living-improved-despite-more-expensive-groceries

Use APA referencing throughout your document.

